---
title: "Economical and health consequences of severe weather events"
author: "Bob de Theije"
date: "October 30, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
The basic goal of this study is to explore the NOAA Storm Database and answer the following questions about severe weather events:

1. Across the United States, which types of events are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing
The data come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size.

### Loading librarys needed
```{r,warning=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
library(reshape2)
```
### download file and read data
```{r cache = TRUE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
# download file if not already in Wd
setwd("c:/Users/Bob/Documents/coursera/data")

if(!file.exists("stormdata.bz2"))
{
  download.file(fileUrl, "stormdata.bz2")
}

# read if necessary
if (!exists("storm_data"))
{
    storm_data<-read.csv("stormdata.bz2", head=TRUE)
}
```
### subsetting the data
only select the attributes relevant to the questions to be answered.

1. which types of events are most harmful with respect to population health?

EVTYPE

FATALITIES

INJURIES

2. which types of events have the greatest economic consequences?

EVTYPE

PROPDMG

PROPDMGEXP

CROPDMG

CROPDMGEXP

  
```{r}
sub_storm_data <- select(storm_data, c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP"))
```

### check for null values
```{r}
if (mean(is.na(sub_storm_data$EVTYPE)) + mean(is.na(sub_storm_data$FATALITIES)) + mean(is.na(sub_storm_data$INJURIES)) + mean(is.na(sub_storm_data$PROPDMG)) + mean(is.na(sub_storm_data$PROPDMGEXP)) + mean(is.na(sub_storm_data$CROPDMG)) + mean(is.na(sub_storm_data$CROPDMGEXP)) > 0)
    print("missing values present") else print("no missing values")
```

### Check the event types
```{r}
EVTYPEC <- count(sub_storm_data, EVTYPE)
arrange(EVTYPEC, desc(n))
```

### cluster events
We see that there are events that look the same
f.i. TSTM WIND, THUNDERSTORM WIND, THUNDERSTORM WINDS and HIGH WIND
or all kinds of FLOOD
```{r}
sub_storm_data <- mutate(sub_storm_data, 
                         EVENT = case_when (str_detect(str_to_lower(EVTYPE), "wind")   ~ "WIND",
                                            str_detect(str_to_lower(EVTYPE), "flood")  ~ "FLOOD",
                                            TRUE ~ as.character(EVTYPE)
                         ))


```

### check the units of dollars in de EXP columns
```{r}
sort(table(sub_storm_data$PROPDMGEXP), decreasing = TRUE)
sort(table(sub_storm_data$CROPDMGEXP), decreasing = TRUE)
```
There is some mess in units, so we transform those variables in one unit (dollar) variable by the following rule:

* K or k: thousand dollars (10^3)

* M or m: million dollars (10^6)

* B or b: billion dollars (10^9)

the rest would be consider as dollars

```{r}
sub_storm_data <- mutate(sub_storm_data, 
                         PROPDAMAGE = case_when (str_detect(str_to_lower(PROPDMGEXP), "k") ~ PROPDMG*1000 ,
                                            str_detect(str_to_lower(PROPDMGEXP), "m")      ~ PROPDMG*1000000,
                                            str_detect(str_to_lower(PROPDMGEXP), "b")      ~ PROPDMG*1000000000,
                                            TRUE ~ PROPDMG
                         ))


sub_storm_data <- mutate(sub_storm_data, 
                         CROPDAMAGE = case_when (str_detect(str_to_lower(CROPDMGEXP), "k") ~ CROPDMG*1000 ,
                                            str_detect(str_to_lower(CROPDMGEXP), "m")      ~ CROPDMG*1000000,
                                            str_detect(str_to_lower(CROPDMGEXP), "b")      ~ CROPDMG*1000000000,
                                            TRUE ~ CROPDMG
                         ))
sub_storm_data <- mutate(sub_storm_data, TOTDAMAGE = PROPDAMAGE + CROPDAMAGE)
```

check to see if the mutation makes sense
```{r}
head(sub_storm_data[sub_storm_data$PROPDMG > 0,c("PROPDMG", "PROPDMGEXP", "PROPDAMAGE")])
head(sub_storm_data[sub_storm_data$CROPDMG > 0,c("CROPDMG", "CROPDMGEXP", "CROPDAMAGE")])
```


# RESULTS

## Across the United States which types of events are most harmfull with respect to population health.

```{r}
event_fatalities_top10 <- group_by(sub_storm_data, EVENT) %>% summarise(sum_fatalities = sum(FATALITIES)) %>% arrange(desc(sum_fatalities)) %>% head(10)
event_injuries_top10 <- group_by(sub_storm_data, EVENT) %>% summarise(sum_injuries = sum(INJURIES)) %>% arrange(desc(sum_injuries)) %>% head(10)
```

### Top 10 Events causing fatalities 

```{r}
print(event_fatalities_top10)
g<- ggplot(event_fatalities_top10) + geom_col(aes(EVENT, sum_fatalities, fill = EVENT))
g <- g + theme(axis.text.x = element_text(angle=45, hjust=1))
print(g)
```

### Top 10 Events causing injuries

```{r}
print(event_injuries_top10)
g<- ggplot(event_injuries_top10) + geom_col(aes(EVENT, sum_injuries, fill = EVENT))
g <- g + theme(axis.text.x = element_text(angle=45, hjust=1))
print(g)
```

It's obvious that Tornado's are most harmfull with respect to population health causing both most fatalities and injuries.


## Across the United States, which types of events have the greatest economic consequences.

```{r}
event_crop_damage_top10 <- group_by(sub_storm_data, EVENT) %>% summarise(crop_damage = sum(CROPDAMAGE)/1000000)  %>% 
                    arrange(desc(crop_damage)) %>% head(10) 

event_property_damage_top10 <- group_by(sub_storm_data, EVENT) %>% summarise(property_damage = sum(PROPDAMAGE)/1000000)  %>% 
                    arrange(desc(property_damage)) %>% head(10) 


event_total_damage_top10 <- group_by(sub_storm_data, EVENT) %>% summarise(total_damage = sum(TOTDAMAGE)/1000000
                                                                    , crop_damage = sum(CROPDAMAGE)/1000000
                                                                    , property_damage = sum(PROPDAMAGE)/1000000)  %>% 
                    arrange(desc(total_damage)) %>% head(10) 


```

###Top 10 crop damage
```{r}
print(event_crop_damage_top10)
```

###Top 10 property damage
```{r}
print(event_property_damage_top10)
```

###Top 10 total damage
```{r}
print(event_total_damage_top10)
```


###Top 10 total damage in figures
```{r}
event_total_damage <- melt(event_total_damage_top10, id.vars="EVENT")
g <- ggplot(event_total_damage, aes(x=reorder(EVENT, -value), y=value))
g <- g + geom_bar(stat="identity", aes(fill=variable), position=position_dodge())
g <- g + ylab("Damage in millions of Dollars") 
g <- g + xlab("Event Type") 
g <- g + theme(axis.text.x = element_text(angle=45, hjust=1))
g <- g + ggtitle("Top 10 eventypes for economical damage")
print(g)

```

We see that total economical damage is greatest for flood, mainly due to property damage.

##Conclusion

1. which types of events are most harmful with respect to population health?

    Tornado's are the events which are most harmful, causing the most fatalities and injuries.

2. which types of events have the greatest economic consequences?

    Flood has the greatest economical consequences, mostly due to porperty damage.
    For crop damage drougth had the greatest impact.
    
    
    
    
--