---
title: "Economical and health consequences of severe weather events"
author: "Bob de Theije"
date: "August 1, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
The basic goal of this study is to explore the NOAA Storm Database and answer the following questions about severe weather events:
1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing
The data come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size.

### Loading librarys needed
```{r,warning=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
```
### download file and read data
```{r cache = TRUE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
# download file if not already in Wd
if(!file.exists("repdata%2Fdata%2FStormData.csv.bz2"))
{
  download.file(fileUrl, "stormdata.bz2")
}

# read if necessary
if (!exists("storm_data"))
{
    storm_data<-read.csv("stormdata.bz2", head=TRUE)
}
```
### subsetting the data
only select the attributes relevant to the questions to be answered.
1. which types of events are most harmful with respect to population health?
EVTYPE
FATALITIES
INJURIES
2. which types of events have the greatest economic consequences?
EVTYPE
PROPDMG
PROPDMGEXP
CROPDMG
CROPDMGEXP
  
```{r}
sub_storm_data <- select(storm_data, c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP"))
```

### check for null values
```{r}
if (mean(is.na(sub_storm_data$EVTYPE)) + mean(is.na(sub_storm_data$FATALITIES)) + mean(is.na(sub_storm_data$INJURIES)) + mean(is.na(sub_storm_data$PROPDMG)) + mean(is.na(sub_storm_data$PROPDMGEXP)) + mean(is.na(sub_storm_data$CROPDMG)) + mean(is.na(sub_storm_data$CROPDMGEXP)) > 0)
    print("missing values present") else print("no missing values")
```

### Check the event types
```{r}
EVTYPEC <- count(sub_storm_data, EVTYPE)
arrange(EVTYPEC, desc(n))
```

### cluster events
We see that there are events that look the same
f.i. TSTM WIND, THUNDERSTORM WIND, THUNDERSTORM WINDS and HIGH WIND
or all kinds of FLOOD
```{r}
sub_storm_data <- mutate(sub_storm_data, 
                         EVENT = case_when (str_detect(str_to_lower(EVTYPE), "wind")                                                  ~ "WIND",
                                            str_detect(str_to_lower(EVTYPE), "flood")                                                  ~ "FLOOD",
                                            TRUE ~ as.character(EVTYPE)
                         ))


```

